An overview on temporal databases.\\ New approaches and proposals.

### 2012-02-25 book/sources/timeDomain.tex
update for the graphs/allen.eps

%
% Time domain
%
In order to define and to work with time it is necessary to study and understand the underlying domain. Moreover, the definition of a temporal domain is the basis for almost any temporal system. In the following subsection \ref{subsec:basic-concepts} we will introduce several concepts that are widely used in the community of temporal databases. Most of the concepts explained here have been introduced in the \emph{'Consensus Glossary of Temporal Database Concepts'}~\cite{Dyreson:1994:CGT:181550.181560}.

\subsection{Basic concepts and properties}
\label{subsec:basic-concepts}

Time is usually represented as a set with an imposed partial order. There are two main models of time: \textbf{linear} and \textbf{cyclic}. In the linear model, time advances from the past to the future in a totally ordered fashion (a relationship of total order is imposed to the set). A cyclic model of time is used for recurrent processes. The majority of the proposals work with a linear model of time.

\begin{svgraybox}
The basic elements in a temporal domain are the following:\\
%\begin{itemize}
%\item
\textbf{Instant}:  Is a time point over an underlying temporal domain.\\
%\item
\textbf{Time interval}: Is the time between two instants.\\
%\item
\textbf{Event}: Is an instantaneous fact. Something that happens at an instant.\\
%\item
\textbf{Chronon}: Is a non-decomposable time interval of fixed duration. It is possible for a temporal model to leave the chronon duration unspecified. The duration will be specified later in the implementation of the model.
%\end{itemize}
\end{svgraybox}
Any temporal model may represent the time as points as well as intervals. The equivalence between a interval-based and point-based temporal data models is demonstrated in \cite{Böhlen_point-versus}.

Attending to the density between the chronons, a model can be classified into the following three types:

\begin{itemize}
\item
\textbf{Discrete model}:  The discrete model of time is isomorphic in relation to the natural numbers. According to this model, each natural number corresponds to a \emph{chronon}. 
\item
\textbf{Dense model}: This model is isomorphic in relation to the rational or the real numbers. It represents one instant of time in the gap between another two instants.
\item
\textbf{Continuous model}: This is an extension of the dense model with no gaps. It is isomorphic in relation to the real numbers.
\end{itemize}

Restrictions on time range may be performed on the base of the following two concepts:
\begin{itemize}
\item
\textbf{Time boundedness}: Time can be bounded orthogonally in the past and in the future.
\item
\textbf{Time distance}: As a metric, time has a distance function which presents four properties:
\begin{enumerate}
\item
Distance is non-negative.
\item
The distance between two different elements is not equal to zero.
\item
The distance from $\alpha$ to $\beta$ is the same as that between $\beta$ and $\alpha$.
\item
Triangle inequality: The distance from $\alpha$ to $\gamma$ is equal or greater than the distance from $\alpha$ to $\beta$ plus the distance from $\beta$ to $\gamma$.
\end{enumerate}
\end{itemize}

On the other hand, time may be \textbf{relative} or \textbf{absolute} (in other words, \textbf{anchored} or \textbf{unanchored}). Sometimes, what one considers \emph{absolute time} is not as definite as one would hope since our concept of absolute time is based on another time reference (for example, January 1st of year one). Relative time has a direction, differently from distance. It is also possible to use negative references so as to represent relative time (e.g. -3 days = three days ago).

%%% granularity
%\subsubsection{Granularity}
Next to the inherent difficulties of time, an issue is \textbf{temporal granularity}. A temporal granularity is a partitioning of the time line used by a system, usually dependent on the application. For example, the age of an adult human being is usually expressed in years: one will use sentences like \emph{`Laura is 21 years old'} instead of \emph{`Laura is 21 years, 3 months and 4 days old'}. In this example any time period shorter than a year needs no representation and thus the used granularity allows no specification for time periods shorter than a year.

The transition between granularities~\cite{Lin97} is also considered as a source of imprecision. In a sentence like \emph{`The Alhambra is 612 years old'}, the temporal notion is very precise if time periods shorter than a year are irrelevant in the used granularity. However, if one is concerned with the exact day on which the last stone of the building is laid, a transition to a finer granularity, allowing the specification of time periods shorter than a year, is needed. Therefore, some proposals consider granularity as the base of the temporal model~\cite{Cru97}.

\subsection{Temporal Relations}
%an explanation on allen's temporal relations.
%definition of fuzzy temporal interval

Several operators are defined in order to compare temporal elements. Allen~\cite{Allen83} first described the relations between time intervals and to a lesser extent, between time points. Figure \ref{fig:allen} shows the temporal relations. There are some proposals that can be applied to both crisp and fuzzy temporal interval ~\cite{ohlbach2004},\cite{nagypal2003},\cite{schockaert08}. For this fuzzy comparisons~\cite{garrido2009} proposed some different temporal relations that can be easyly implemented by means of regular fuzzy operations.

\begin{figure}
\centering
\input{graphs/allen.tex}
\caption{Allen relations between two time intervals.}
\label{fig:allen}
\end{figure}

\subsection{An example of temporal domain: Julian Day Number}
% proposal for the underlying Julian Day Number domain.

The Julian Day number \emph{JDN} \cite{Dir96} is a counter. Its value is incremented in one unit every day from 1 January 4713 B.C. at 12:00 noon. The particularity of starting at noon was useful for astronomers: the observations they took one night belonged to the same Julian Day.\\
Note that the Julian Day represents whole days. There is an extension that allows to represent any precision needed (it is called Julian Date). By default, a Julian Day is expressed in Universal Time. (U.T, also known as Solar Time). However, there are representations in Terrestrial Time (T.T), Epheris Time (J.E.D. or J.D.E.). Any time scale different from Universal Time must be explicited after the Julian Day number. There are several conversion formulas~\cite{Usn}\cite{Wik}\cite{Lea} between a date in Gregorian calendar  format and a date in  JDN format. The inverse conversion formula is proposed in~\cite{Fliegel:1968:LEM:364096.364097}\\

@@

%\subsubsection{Conversion among dates}
%\paragraph{Conversion between Gregorian and Julian dates}
%The Julian Date consist on the Julian Day number and the time. It is computed with the formula \cite{Usn}:\\
%\begin{equation}
%\label{eq:julian day}
%JD = JDN + \frac{hour-12}{24} + \frac{minutes}{1440} + \frac{seconds}{86400}
%\end{equation}
%
%Here the JDN is the Julian Day Number. This equation does not take into account the leap seconds \footnote{A leap second(s) is the amount of second(s) added or deleted every year in the Universal Time.\cite{Wik},\cite{Lea}. }.

%\paragraph{Gregorian calendar to Julian day}
%Any date from Gregorian calendar \footnote{As Gregorian calendar entered in use in 1582, any use before that date is called \emph{proleptic calendar}.} can be converted into a Julian Day using the following formula:\\
%\begin{equation}
%\label{eq:JDN}
%JDN = \frac{1461x(Y + 4800 + \frac{M-14}{12})}{4}+\frac{367x(M-2-12x\frac{M-14}{12})}{12}-\frac{3x\frac{Y+4900+\frac{M-14}{12}}{100}}{4} 
%\end{equation}
%
%This formula was created by Henry Fliegel and Thomas C. van Flandern in 1968. \cite{Fliegel:1968:LEM:364096.364097} It takes into account leap years and that they are applicable only when the century is divisible by 400. The value $Y$ represents the year and the value $M$ represents the month.
%\paragraph{Julian calendar to Julian day}
%The algorithm to convert between Julian calendar and Julian Day Number is only valid for dates since 4713 B.C (positive values of Julian Day) \cite{Dog}:\\
%\begin{equation}
%\label{eq:JDN_2_JD}
%JDN=367Y-\frac{7 (Y + 5001 + \frac{M-9}{7})}{4}+\frac{275xM}{9}+D+1729777
%\end{equation}
%
%Where $Y$ is the year value, $M$ is the month value, and $D$ is the day value.
%
%%\subsubsection{Membership degree computation}
%The main advantage from the Julian day number is the ability to make calculations. Thus, the membership degree computation for a fuzzy number based on a Julian day is explained here. \\
%Let a trapezoidal possibility distribution defined by four JDN named $a$,$b$,$c$ and $d$. The fuzzy validity period \emph{FVP}, can be represented as in figure \ref{fig:fuzzy-validity-period}. The following equation shows the membership degree for a Julian day number \emph{JDN} in a \emph{FVP} represented by means of the possibility distribution of $[a,b,c,d]$. In a valid-time database, the values for the valid starting time \emph{VST} and the valid ending time \emph{VET} can be fuzzyfied as shown in equation \ref{eq:fuzzy-validity-period}.
%
%The membership function for a given $x$ number in \emph{JDN} format is the following:\\
%\begin{equation}
%\label{eq:trapezoidal distribution}
%%\[
%T(x)=
%\begin{cases}
%0, \mbox(if x \leq a, x \geq d)\\
%1, \mbox(if x \in [b,c])\\
%\frac{x-a}{b-a}   \mbox(if x \in (a,b])\\
%\frac{d-x}{d-c} 	 \mbox(if x \in (c,d])\\
%\end{cases}
%%\]
%\end{equation}
%The points outside the trapezium or inside the interval $[a,b]$ are trivial. The points inside the other intervals has the following particularities:\\
%The subtraction operation with Julian day numbers gives as result a scalar value (as explained before). The division operation between two scalars gives as result an scalar in the interval $[0,1]$ which represents the membership degree.

>>>>

%
% Time domain
%
Before considering the introduction of temporal modelling to information systems, it is necessary to define and explain some main concepts concerning temporal modelling and their corresponding terminology, to situate these and to discuss some properties and issues related to these concepts. In this section, several basic concepts and corresponding terminology will be defined, explained and situated. Most of these concepts are widely used in the community of temporal databases and their definitions have been agreed upon in the context of~\cite{Dyreson1994}. For these concepts, in the entire chapter, the contents of~\cite{Dyreson1994} are followed (and often cited).

In subsection \ref{subsec:basic-concepts}, some basic concepts of time modelling are presented and explained.

%In order to define and to work with time it is necessary to study and understand the underlying domain. Moreover, the definition of a temporal domain is the basis for almost any temporal system. In the following subsection \ref{subsec:basic-concepts} we will introduce several concepts that are widely used in the community of temporal databases. Most of the concepts explained here have been introduced in the \emph{'Consensus Glossary of Temporal Database Concepts'}~\cite{Dyreson1994}.

\subsection{Basic concepts and properties}
\label{subsec:basic-concepts}
In information systems, time itself is usually perceived as a linear or cyclic concept~\cite{Jensen94thetsql2}. Therefore, a time domains modelling time is usually represented by a set with an imposed partial order. In general, two main types of time models can be discerned: a \emph{linear} model~\cite{benthem82} and a \emph{cyclic} model~\cite{lorentzos88}. In the linear model, a total order is imposed on the set and the progress of time is seen as a linear matter, while cyclic models are mainly used in the modelling of recurrent processes. It should be noted that the majority of proposals use linear time models.

Data models used by information systems (and in specific, temporal database systems) may represent these underlying time domains using \emph{chronons}~\cite{Dyreson1994}. However, to explain what chronons are, an explanation of some other temporal concepts is necessary.

\begin{svgraybox}
\vspace{-10pt}
\begin{definition}\textbf{Instant}~\cite{Dyreson1994}\\
An \emph{\textbf{instant}} is a time point on an underlying time axis.
\end{definition}
\vspace{-10pt}
\end{svgraybox}

Thus, an instant is basically an instantaneous time point in a time model.

Orthogonal to the classification of underlying time models as linear or cyclic, they can be classified as \emph{discrete}, \emph{dense} or \emph{continuous} models. In a discrete model~\cite{Clifford:1985:AHR:971699.318922}, the notion exists that every instant has a unique successor and the set of instants is seen as a discrete one. Here, intuitively, the set of instants can be seen as isomorphic to the set of natural numbers $\Nat$. In a dense model, the notion exists that between any two instants always lies another, while the set of instants can still be seen as a discrete one. Here, intuitively, the set of instants can be seen as isomporphic to the set of rational numbers $\Q$ or the set of real numbers $\R$. In a continuous model, the notion also exists that between any two instants always lies another one, but unlike dense models, the set of instants is seen as continuous and there are no ``gaps'' between successive instants.

Some other necessary concepts are ~\cite{Dyreson1994}:

\begin{svgraybox}
\vspace{-10pt}
\begin{definition}\textbf{Time Interval}\\
A \emph{\textbf{time interval}} is the time between two instants.
\end{definition}

\begin{definition}\textbf{Temporal Element}\\
A \emph{\textbf{temporal element}} is a finite union of time intervals.
\end{definition}

\begin{definition}\textbf{Event}\\
An \emph{\textbf{event}} is an instantaneous fact, i.e. something occurring at an instant.
\end{definition}

\begin{definition}\textbf{Duration}\\
A \emph{\textbf{duration}} is an amount of time with known length, but no specific starting or ending instants.
\end{definition}

\begin{definition}\textbf{`Temporal' as Modifier}\\
The modifier \emph{\textbf{`temporal'}} is used to indicate that the modified concept concerns some aspect of time.
\end{definition}
\vspace{-10pt}
\end{svgraybox}

%In the linear model, time advances from the past to the future in a totally ordered fashion (a relationship of total order is imposed to the set). A cyclic model of time is used for recurrent processes. The majority of the proposals work with a linear model of time.

Data models might now represent an underlying time domain using chronons:

\begin{svgraybox}
\vspace{-10pt}
\begin{definition}\textbf{Chronon}\\
In a data model, a \emph{\textbf{chronon}} is a non-decomposable time interval of some fixed, minimal duration.
\end{definition}
\vspace{-10pt}
\end{svgraybox}

A data model may now represent a time model or time domain by a sequence of consecutive chronons. These chronons have identical durations. A data model will typically not specify the exact chronon duration, so it can be fixed later by applications implementing the data model.

The fact that chronons are actually time intervals has a particular effect on the representation of instants and time intervals. In a data model representing a time domain using chronons, an instant is of course represented as a chronon, but one single chronon may represent multiple instants. A time interval may be represented by a set of contiguous chronons, depending on the amount of time the time interval comprises.

%\begin{svgraybox}
%The basic elements in a temporal domain are the following:\\
%\begin{itemize}
%\item
%\textbf{Instant}:  Is a time point over an underlying temporal domain.\\
%\item
%\textbf{Time interval}: Is the time between two instants.\\
%\item
%\textbf{Event}: Is an instantaneous fact. Something that happens at an instant.\\
%\item
%\textbf{Chronon}: Is a non-decomposable time interval of fixed duration. It is possible for a temporal model to leave the chronon duration unspecified. The duration will be specified later in the implementation of the model.
%\end{itemize}
%\end{svgraybox}

%\begin{definition}\textbf{(set of chronons)}
%\label{def:set-chronon}
%The \textbf{\emph{set of all the chronons}} in a system is denoted by $\C$. A chronon $c$ is therefore defined as:
%\begin{equation}
%\left \lbrace c \mid c \in \C \right \rbrace
%\end{equation}
%\end{definition}

Another classification of time models concerns the use of points or intervals to model time. The equivalence between interval-based and point-based time models is demonstrated in \cite{Böhlen_point-versus}.

%Attending to the density in the set $\C$ of the chronons, a model can be classified into the following three types:

%\begin{itemize}
%\item
%\textbf{Discrete model} ($\C \subset \Nat$):  The discrete model of time is isomorphic in relation to the natural numbers. %According to this model, each natural number corresponds to a \emph{chronon}.
%\item
%\textbf{Dense model} ($\C \subset \Q$ or $\C \subset \R$): This model is isomorphic in relation to the rational or the real %numbers. It represents one instant of time in the gap between another two instants. 
%\item
%\textbf{Continuous model} ($\C \subset \R$): This is an extension of the dense model with no gaps. It is isomorphic in relation to the real numbers. 
%\end{itemize}

\textcolor{red}{CHECK}
Restrictions on time range may be performed on the base of the following two concepts~\cite{Jensen94thetsql2}:
\begin{itemize}
\item
\textbf{Time boundedness}: Time can be bounded orthogonally in the past and in the future.
\item
\textbf{Time distance}: As a metric, time has a distance function which presents four properties:
\begin{enumerate}
\item
Distance is non-negative.
\item
The distance between two different elements is not equal to zero.
\item
The distance from $\alpha$ to $\beta$ is the same as that between $\beta$ and $\alpha$.
\item
Triangle inequality: The distance from $\alpha$ to $\gamma$ is equal or greater than the distance from $\alpha$ to $\beta$ plus the distance from $\beta$ to $\gamma$.
\end{enumerate}
\end{itemize}
\textcolor{red}{END CHECK}

\textcolor{red}{MOVE TO TEMPORAL DATABASE PART}
On the other hand, time may be~\cite{Jensen94thetsql2} \textbf{relative} or \textbf{absolute} (in other words, \textbf{anchored} or \textbf{unanchored}). Sometimes, what one considers \emph{absolute time} is not as definite as one would hope since our concept of absolute time is based on another time reference (for example, January 1st of year one). Relative time has a direction, differently from distance. It is also possible to use negative references so as to represent relative time (e.g. -3 days = three days ago).
\textcolor{red}{END MOVE}

Subsection \ref{subsec:granularity} gives a short survey on the concept of temporal granularities.

%%% granularity
\subsection{Granularities}\label{subsec:granularity}
An important issue in time modelling concerns the concept of \emph{temporal granularities}. A formal definition for this concept is given in~\cite{Lin97}:

\begin{svgraybox}
\vspace{-10pt}
\begin{definition}\textbf{Granularity}~\cite{wang93}\\
A \emph{\textbf{granularity}} is an ordered set of non-overlapping and continuous temporal elements called \emph{granules}.
\end{definition}

\begin{definition}\textbf{Granule}\\
A \textbf{\emph{granule}} is the basic time unit in a granularity.
\end{definition}
\vspace{-10pt}
\end{svgraybox}

A temporal granularity is in fact a partitioning of the time line (time model) used by a system, usually dependent on the application. For example, the age of an adult human being is usually expressed in years: one will use sentences like `Laura is 21 years old' instead of sentences like `Laura is 21 years, 3 months and 4 days old'. In this example any duration shorter than a year needs no representation and thus the used granularity allows no specification for durations shorter than a year. The granules are years.

As a granularity $G$ is an ordered set, each granule may be represented by an integer. In this representation, to keep track of the granularity a granule is an element of, the corresponding granularity is added in subscript:

\begin{equation}
G = \left\{i_{G}\ |\ i \in \Z \right\}
\end{equation}

%Next to the inherent difficulties of time, an issue is \textbf{temporal granularity}. A temporal granularity is a partitioning of the time line used by a system, usually dependent on the application. For example, the age of an adult human being is usually expressed in years: one will use sentences like \emph{`Laura is 21 years old'} instead of \emph{`Laura is 21 years, 3 months and 4 days old'}. In this example any time period shorter than a year needs no representation and thus the used granularity allows no specification for time periods shorter than a year.
%The formal definition for granularity is given in ~\cite{Lin97}:

%\begin{definition}
%\label{def:granularity}\textbf{(granularity)}
%A \textbf{\emph{granularity}} $\alpha$ is an ordered set of non-overlapping and continuous time elements called granules. \\
%\end{definition}

%As granularity $\alpha$ it is an ordered set, each \emph{granule} may be indexed by an integer number:\\
%\begin{equation}
%\alpha = \left \lbrace \ldots, 0_\alpha, 1_\alpha, \ldots \right \rbrace
%\end{equation}

In a system, the granularity with the shortest granules is the \emph{chronons granularity}, which is denoted as $\bot$. It is the granularity of which the granules are chronons. 
%It is now possible to define a function $f$, called a \emph{mapping function}, that maps a given granule $i_G$, $i \in \Z$, from a given granularity $G$ to a set of corresponding chronons. The function definition should look like this:

\begin{svgraybox}
\begin{definition}
\label{def:mapping-function}
\textbf{Mapping function}~\cite{Lin97}\\ 
A mapping function $f$ is a function that translates from a given granule $i_G$, $i \in \Z$, in a given granularity $G$, to a set of corresponding chronons.\\

\vspace{-10pt}

\begin{align}
f & : G \rightarrow \Pow(\bot)\nonumber\\
  & : i_{G} \rightarrow \lbrace c_\bot\ |\ (c_\bot\text{ is contained in }i_{G}) \wedge (c_\bot \in \bot)\rbrace\nonumber
\end{align}
\end{definition}
\end{svgraybox}

Note that a mapping function $f$ always maps from a given granularity $G$ to the set of chronons $\bot$. Therefore, the output for a mapping function is a subset of $\Pow(\bot)$. \\
 A mapping function $f$ requires that the following properties hold~\cite{Lin97}:

%
%
%For a given granularity $G$ and a given mapping function $f$ from $G$ to $\bot$, it is required that 
\begin{property}
%\item
$G$ \emph{is an ordered set}: for a given $i_{G} \in G$ holds: if $c_\bot \in f(i_G)$ and $c^{'}_\bot \in f(j_G)$, then $c_\bot < c^{'}_\bot$ implies that $i_G < j_G$. This is illustrated in figure \ref{fig:granularity-prop1}.
\begin{figure}
\centering
\input{graphs/granularity-prop1.tex}
\caption{$G$ is an ordered set}
\label{fig:granularity-prop1}
\end{figure}
\end{property}

%\item
\begin{property}
$G$ \emph{is a set of continuous granules}: for a given $i_{G} \in G$ and given chronons $c_\bot$, $c^{'}_\bot$ and $c^{''}_\bot$ holds: if $c_\bot < c^{'}_\bot < c^{''}_\bot$, then $c_\bot \in f(i_G)$ and $c^{''}_\bot \in f(i_G)$ implies that $c^{'}_\bot \in f(i_G)$. This is illustrated in figure \ref{fig:granularity-prop2}.
\begin{figure}
\centering
\input{graphs/granularity-prop2.tex}
\caption{$G$ is a set of continuous granules}
\label{fig:granularity-prop2}
\end{figure}
\end{property}

%\item
\begin{property}
\emph{The granules in} $G$ \emph{do not overlap}: for given $i_{G}\text{ and } j_{G} \in G$ holds: if $i_G \neq j_G$, then $f(i_G) \cap f(j_G) = \emptyset$. This is illustrated in figure \ref{fig:granularity-prop3}.
\begin{figure}
\centering
\input{graphs/granularity-prop3.tex}
\caption{The granules in $G$ do not overlap}
\label{fig:granularity-prop3}
\end{figure}
\end{property}
%\end{itemize}

Mapping functions between granularities and the chronons granularity also allow comparing granularities with respect to the length of their granules. In this context, two important concepts can be discerned.
%A granularity $G$ is said to be \emph{\textbf{finner than}} a granularity $H$ when the set of chronons on which a given granule $i_G$ is mapped by a mapping function $f$ has less elements than the set of chronons on which a given granule $j_H$ is mapped by $f$.
%A granularity $\alpha$ is \emph{``coarser than''} granularity $\beta$ when the set of corresponding chronons for a given granule $i_\alpha$ is greater than the set of the corresponding chronons for a given granule $j_\beta$.

\begin{svgraybox}
\vspace{-10pt}
\begin{definition}\textbf{Finner Than}\\
Consider a mapping function $f$ and let $i_G$ and $j_H$ be elements of granularities $G$ and $H$ respectively. Granularity $G$ is now said to be \emph{\textbf{finner than}} granularity $H$ if:
\begin{equation}
\label{eq:finner-than}
| f \left( i_G \right) |\ <\ |  f \left( j_H \right)| \nonumber
\end{equation} 
\end{definition}

\begin{definition}
\label{def:coarser-than}\textbf{Coarser Than}
Consider a mapping function $f$ and let $i_G$ and $j_H$ be elements of granularities $G$ and $H$ respectively. Granularity $G$ is now said to be \emph{\textbf{coarser than}} granularity $H$ if:
\begin{equation}
\label{eq:coarser-than}
| f \left( i_G \right)|\ >\ |  f \left( j_H \right)| \nonumber
\end{equation} 
\end{definition}
\vspace{-10pt}
\end{svgraybox}

It is also possible to describe the relation between different granularities. This is called a casting function:
\begin{svgraybox}
\begin{definition}
\textbf{Casting function}~\cite{Lin97}\\
\label{def:casting-fucntion}
A casting function $cast$ maps granules from a granularity $G$ to granules from a different granularity $H$.
\begin{equation}
\label{eq:cast-function}
\mbox{cast} \left( i_G,G,H \right) = j_H \nonumber
\end{equation}
\end{definition}
\end{svgraybox}

This notation indicates that the function $cast$ associates a granule $i_G$ in $G$ to a corresponding granule $j_H$ in $H$. Two kinds of granularity-to-granularity mappings can now be discerned: an \emph{upwards mapping} is a mapping from a granularity $G$ to a coarser granularity $H$, whereas a downwards mapping is a mapping from a granularity $K$ to a finner granularity $L$. Different granularity-to-granularity mappings between several granularities can be represented using a granularity graph, which is a directed graph indicating the mapping conversions.

 Mappings between two granularities may be classified as \emph{regular mappings}, \emph{irregular mappings} or \emph{congruent mappings}.
\begin{itemize}
\item
\emph{Regular mapping}: A regular mapping is a granularity-to-granularity mapping where the mapping function between the two granularities is calculated by means of multiplications and/or divisions and (maybe) an anchor adjustment. E.g. the mapping between hours and minutes is calculated using a multiplication by $60$. 
\item
\emph{Irregular mapping}: An irregular mapping is a granularity-to-granularity mapping where the mapping function can not be calculated by means of multiplications and/or divisions. E.g. the mapping between months and days is dependent on the exact month or day. 
\item
\emph{Congruent mapping}: A congruent mapping is a granularity-to-granularity mapping where the two granularities in the mapping have the same granules but a different anchor. E.g. the mapping between the days (Gregorian calendar days) and the academic days. 
\end{itemize}

\begin{example}
Consider a systems that models both Gregorian Calendar dates as well as Academic dates. The chronons granularity in this system are milliseconds. Figure \ref{fig:granularity-graph-example} shows the complete granularity graph example. The transitions between Chronons and Seconds and between Seconds and Minutes and so on is called a \emph{regular mapping}. The transition between Days and Months is not defined by a multiplication or division, this is called a \emph{irregular mapping}. In the graph, these \emph{irregular mappings} are noted with a bold arrow. Finally, the transition between a Gregorian Calendar Day and an Academic Day is called a \emph{Congruent mapping}: two granularities with the same granules but with different anchors. Both concern the same days, but the academic year starts on October 1st, whereas the Gregorian calendar year starts on January 1st. Congruent mappings are visualised as straight lines without arrow heads.

\end{example}

\textcolor{red}{MOVE TO CHAPTER 4}
The transition between granularities~\cite{Lin97} is also considered as a source of imprecision. In a sentence like \emph{`The Alhambra is 612 years old'}, the temporal notion is very precise if time periods shorter than a year are irrelevant in the used granularity. However, if one is concerned with the exact day on which the last stone of the building is laid, a transition to a finer granularity, allowing the specification of time periods shorter than a year, is needed. Therefore, some proposals consider granularity as the base of the temporal model~\cite{Cru97}.

\textcolor{red}{END MOVE TO CHAPTER 4}

\begin{figure}
\centering
\includegraphics[scale=0.5]{graphs/granularityGraph.eps}
\caption{A granularity graph example}
\label{fig:granularity-graph-example}
\end{figure}

Next, in subsection \ref{subsec:timedomain-calendar}, some basic concepts and issues concerning calendars are explained.

\subsection{Calendars}
\label{subsec:timedomain-calendar}
%maybe introduce previously the concept of time units, linear hierarchy and time point.

This subsection will concern \emph{calendars}. A calendar provides some sort of human interpretation of time. As such, calendars provide meaning or interpretation to temporal values where this meaning or interpretation is relevant or useful to the user. More precisely, calendars determine the mapping between human-meaningful time values and an underlying time model~\cite{Dyreson1994}. More formally, a calendar is an organization of different granularities (e.g. granularities day, month and year) in a hierarchy. In order to formally define what a calendar is, some other concepts should first be introduced~\cite{Kraus1997}:

%\begin{svgraybox}
%\vspace{-10pt}
%\begin{definition}\textbf{(Calendar)}\\
%A \emph{\textbf{calendar}} 
%\end{definition}
%\vspace{-10pt}
%\end{svgraybox}

\begin{svgraybox}
\vspace{-10pt}
\begin{definition}\textbf{Linear Hierarchy of Granularities}\\
A \emph{\textbf{linear hierarchy of granularities}} is a finite set of granularities with a linear order (denoted $\sqsubset$). The most coarse granularity is called \emph{the top of the hierarchy} and must be an infinite set while all other granularities must be finite sets.
\end{definition}
\vspace{-10pt}
\end{svgraybox}

An example of a linear hierarchy is the set $\left\{ \text{days, months, years} \right\}$, where days $\sqsubset$ months and months $\sqsubset$ years. With this, a \emph{linear calendar} can be defined.

\begin{svgraybox}
\vspace{-10pt}
\begin{definition}\textbf{Linear Calendar}\\
A \emph{\textbf{linear calendar}} consist of a linear hierarchy of granularities and a validity predicate which specifies if a granule is valid in the calendar.
\end{definition}
\vspace{-10pt}
\end{svgraybox}

An example of a linear calendar could be the linear hierarchy $\left\{ \text{days, months, years} \right\}$, where days $\sqsubset$ months and months $\sqsubset$ years, together with a validity predicate that excludes time indications like `32 April 2012' or `30 February 2012'.

%The linear hierarchy days $\sqsubset$ months $\sqsubset$ years represents a linear calendar. It is necessary to define the validity predicates because 17 January 2012 is a valid element of the calendar whereas 30 February 2012 is not.

\textcolor{red}{CHECK}
\subsubsection{\label{subsubsubsec:julian-day-number}An example of temporal domain: Julian Day Number}
% proposal for the underlying Julian Day Number domain.

The Julian Day number \emph{JDN} \cite{Dir96} is a counter. Its value is incremented in one unit every day from 1 January 4713 B.C. at 12:00 noon. The particularity of starting at noon was useful for astronomers: the observations they took one night belonged to the same Julian Day.\\
Note that the Julian Day represents whole days. There is an extension that allows to represent any precision needed (it is called Julian Date). By default, a Julian Day is expressed in Universal Time. (U.T, also known as Solar Time). However, there are representations in Terrestrial Time (T.T), Epheris Time (J.E.D. or J.D.E.). Any time scale different from Universal Time must be explicit after the Julian Day number. There are several conversion formulas~\cite{Usn}\cite{Wik}\cite{Lea} between a date in Gregorian calendar  format and a date in  JDN format. The inverse conversion formula is proposed in~\cite{Fliegel:1968:LEM:364096.364097}\\

@@

\textcolor{red}{END CHECK}

The next section briefly introduces temporal relationships.

\subsection{Temporal Relationships}
%an explanation on allen's temporal relations.
%definition of fuzzy temporal interval
In this section, a brief introduction can be found, concerning \emph{temporal relationships}, sometimes also called `temporal relations'\cite{Billiet:Pons:Matthe:DeTre:Pons:2011:BipolarFuzzy}. Temporal relationships can be seen as relationships between temporal elements or instants.

\def\JPicScale{0.5}
\begin{figure}[h]
\centering
\includegraphics[scale=0.5]{graphs/allen.eps}
\caption{Allen relations between two time intervals.}
\label{fig:allen}
\end{figure}

Several operators are defined in order to compare temporal elements or time points. Allen~\cite{Allen83} first described the relationships between time intervals and as a special case, between time points. Figure \ref{fig:allen} shows the temporal relationships Allen discerned. Some proposals can be applied to both crisp and fuzzy temporal intervals~\cite{ohlbach2004},~\cite{nagypal2003},~\cite{schockaert08}.

\textcolor{red}{MOVE TO FUZZY PART}
For this fuzzy comparisons~\cite{garrido2009} proposed some different temporal relations that can be easily implemented by means of regular fuzzy operations.
\textcolor{red}{END MOVE TO FUZZY PART}

In the next section, some basic concepts and terminology about temporal databases are presented, followed by an overview of some interesting issues concerning temporal databases and a survey of some commercial temporal database systems.
